Fractional Types

### 2016-12-18 typespp.tex
explicit quantifiers everywhere

Note that for each power $p ^ i$ of $p$, there is a morphism
$\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$ to the
identity. Note also that everything is well-defined even if we choose $p :
\zt\iso\zt$. In that case, the cardinality is 1.

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& [\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& [\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   [\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& [\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  [\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
  & : \epsilon- \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
  & : \epsilon+
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ :[\mathbb{T},v] \leftrightblackspoon
 [\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  [\mathbb{T}_2,v_2] \leftrightblackspoon [\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon [\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 [\mathbb{T}_1,v_1] \leftrightblackspoon
 [\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 [\mathbb{T}_3,v_3] \leftrightblackspoon
 [\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  [\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  [\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
% \\
% \\
% \Rule{}
% {\jdg{}{}{}}
% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
%   \leftrightblackspoon
%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
% {}
\end{array}\]
\\
Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
\forall\exists{k}.\mathbb{T_2}_\bullet$ uses the $\forall$ in the left-to-right
direction and uses the $\exists$ in the right-to-left direction.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We recall that, as all fractional types are inhabited (see remark
in Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going
to pointed types. We then allow universal quantification over indices $k$
associated to values, over a collection of types.  In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
otherwise arbitrary. Once we introduce a universal quantifier in one direction of
execution, we are led to also include an existential quantifier for the reverse
execution. Intuitively, a forward evaluation step that is at liberty to choose a
$k$ will, when reversed encounter a particular, i.e., existentially quantified,
$k$. The syntax is therefore as follows.

@@

@@

\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\longmapsto}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\longmapsto}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\longmapsto}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
%% &\stackrel{\synch_i}{\longmapsto}&
&k\mapsto i&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\exists&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\longmapsto}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\longmapsto}&
  [\order{c}, c^i]
\end{array}\]

\[\begin{array}{rcl}
  [\order{c}, c^i]
&\stackrel{\unitetr/}{\longmapsfrom}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\longmapsfrom}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\assocrt/}{\longmapsfrom}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^k \rangle,c^k \rangle] \\
&k\mapsto i&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\exists&
  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\longmapsfrom}&
  [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
&\stackrel{\unitetl/}{\longmapsfrom}&
  [\order{c},c^i]
\end{array}\]

>>>>

Note that for each power $p ^ i$ of $p$, there is a morphism
$\triple{k}{q}{\alpha}$ in $\iter{p}$ such that $q$ annihilates $p^i$ to the
identity.

\paragraph*{Remark.} Note also that everything is well-defined even if we choose
$p : \zt\iso\zt$. In that case, the cardinality is 1.

@@

\begin{figure}[t]
{\footnotesize
\[\begin{array}{rrcll}
\unitetl/ :& \forall\exists{\overline{k}}.[\order{\idiso} \circledast \mathbb{T},
                                                   \langle \idiso,v \rangle]
  & \leftrightblackspoon &
                   \forall\exists{\overline{k}}.[\mathbb{T},v]
  &: \unititl/ \\
\unitetr/ :& \forall\exists{\overline{k}}.[\mathbb{T} \circledast \order{\idiso},
                                                    \langle v,\idiso]
  & \leftrightblackspoon &
                   \forall\exists{\overline{k}}.[\mathbb{T},v]
  &: \unititr/ \\
\swapt/ :& \forall\exists{\overline{k}}.[\mathbb{T}_1 \circledast \mathbb{T}_2,
                                                   \langle v_1,v_2 \rangle]
 & \leftrightblackspoon &
                  \forall\exists{\overline{k}}.[\mathbb{T}_2 \circledast \mathbb{T}_1,
                                                  \langle v_2,v_1 \rangle]
  &: \swapt/ \\
  \assoclt/ :& \forall\exists{\overline{k}}.
               [\mathbb{T}_1 \circledast (\mathbb{T}_2 \circledast \mathbb{T}_3),
                                        \langle v_1,\langle v_2,v_3 \rangle]
  & \leftrightblackspoon &
                      \forall\exists{\overline{k}}.
               [(\mathbb{T}_1 \circledast \mathbb{T}_2) \circledast \mathbb{T}_3,
                                       \langle \langle v_1,v_2 \rangle , v_3 \rangle]
  &: \assocrt/ \\
\\
\eta- :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\iorder{c} \circledast \order{c}, \langle 1/c^k, c^k  \rangle]
  & : \epsilon- \\
\eta+ :& [\order{\idiso},\idiso] & \leftrightblackspoon &
  \forall\exists{k}.[\order{c} \circledast \iorder{c}, \langle c^k,1/c^k \rangle]
  & : \epsilon+
\end{array}\]
\[\begin{array}{c}
\Rule{}
{}
{\jdg{}{}{\idiso/ : \forall\exists{\overline{k}}.[\mathbb{T},v] \leftrightblackspoon
   \forall\exists{\overline{k}}.[\mathbb{T},v]}}
{}
~~
\Rule{}
{\jdg{}{}{\alpha : c_1 \isotwo c_2}}
{\jdg{}{}{(\order{\alpha})_k :
 [\order{c_1},c_1^k]  \leftrightblackspoon
 [\order{c_2}.c_2^k]}}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
  \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]} \quad
  \rho_2 :
  \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2] \leftrightblackspoon
  \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3]}
{\jdg{}{}{\rho_1 \circledcirc \rho_2 :
 \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
 \forall\exists{\overline{k_3}}.[\mathbb{T}_3},v_3]}
{}
\\
\\
\Rule{}
{\jdg{}{}{\rho_1 :
 \forall\exists{\overline{k_1}}.[\mathbb{T}_1,v_1] \leftrightblackspoon
 \forall\exists{\overline{k_2}}.[\mathbb{T}_2,v_2]}
 \qquad \rho_2 :
 \forall\exists{\overline{k_3}}.[\mathbb{T}_3,v_3] \leftrightblackspoon
 \forall\exists{\overline{k_4}}.[\mathbb{T}_4,v_4]}
{\jdg{}{}{\rho_1 \circledast \rho_2 :
  \forall\exists{\overline{k_1k_3}}.[\mathbb{T}_1 \circledast \mathbb{T}_3, \langle v_1,v_3 \rangle]
  \leftrightblackspoon
  \forall\exists{\overline{k_2k_4}}.[\mathbb{T}_2 \circledast \mathbb{T}_4}, \langle v_2,v_4 \rangle]}
{}
% \\
% \\
% \Rule{}
% {\jdg{}{}{}}
% {\jdg{}{}{\synch_i : \forall{\overline{k}k'}.[\mathbb{T},v]
%   \leftrightblackspoon
%    \forall{\overline{k}}.[\mathbb{T},v[k'\mapsto i]]}}
% {}
\end{array}\]
\\
Each combinator $\rho_\bullet$ has an inverse. A rule $\mathbb{T_1}_\bullet \leftrightblackspoon
\forall\exists{k}.\mathbb{T_2}_\bullet$ uses the $\forall$ in the left-to-right
direction and uses the $\exists$ in the right-to-left direction.
}
\caption{$\Pi^/$ fractional combinators (dependent version)}
\label{pifracd:comb}
\end{figure}

@@

The first step is to enrich the types to pointed types $\mathbb{T_\bullet}$
where each type $\mathbb{T}$ refers to a particular value $v$ in
$\mathbb{T}$. We recall that, as all fractional types are inhabited (see Remark
in Sec.~\ref{sec:symmetryg}), and thus there is no loss in generality in going
to pointed types. We then allow universal quantification over indices $k$
associated to values, over a collection of types.  In particular, the type
$\forall k.[\order{c} \circledast \iorder{c}, \langle c^k, 1/c^k \rangle]$
imposes a constraint on the two values associated with the types $\order{c}$ and
$\iorder{c}$; they must be ``synchronized'' by sharing the same~$k$ but are
otherwise arbitrary. Once we introduce a universal quantifier in one direction
of execution, we are led to also include an existential quantifier for the
reverse execution. Intuitively, a forward evaluation step that is at liberty to
choose a $k$ will, when reversed encounter a particular, i.e., existentially
quantified, $k$. The syntax is therefore as follows.

@@

A type $\mathbb{T}_\bullet$ is generally of the form
$\forall\exists{\overline{k}}.[\mathbb{T},v]$ where all the quantifiers are
lifted to the top leaving a non-dependent type and a value in that type. The
combinators in Fig.~\ref{pifracd:comb} also serve as the operational semantics
for the language as they explain how the value currently in focus changes during
evaluation.

@@

\[\begin{array}{rcl}
[\order{c},c^i]
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{id}\circledast\order{c}, \langle \idiso,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^k, \langle 1/c^k,c^i \rangle\rangle] \\
%% &\stackrel{\synch_i}{\leftrightblackspoon}&
&k\mapsto i&
  [\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^i,c^i \rangle\rangle] \\
&\exists&
  \exists{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}, c^i]
\end{array}\]

\[\begin{array}{rcl}
  [\order{c}, c^i]
&\stackrel{\unitetr/}{\leftrightblackspoon}&
  [\order{c}\circledast\order{\idiso},
  \langle c^i, \idiso \rangle] \\
&\stackrel{\idiso/\circledast\epsilon-}{\leftrightblackspoon}&
  \forall{k}.[\order{c}\circledast(\iorder{c}\circledast\order{c}),
  \langle c^i, \langle 1/c^k,c^k \rangle\rangle] \\
&\stackrel{\assocrt/}{\leftrightblackspoon}&
  \forall{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^k \rangle,c^k \rangle] \\
&k\mapsto i&
  [(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^i,1/c^i \rangle,c^i \rangle] \\
&\exists&
  \exists{k}.[(\order{c}\circledast\iorder{c})\circledast\order{c},
  \langle \langle c^k,1/c^k \rangle,c^i \rangle] \\
&\stackrel{\eta+\circledast\idiso/}{\leftrightblackspoon}&
  [\order{\idiso}\circledast\order{c}, \langle \idiso,c^i\rangle] \\
&\stackrel{\unitetl/}{\leftrightblackspoon}&
  [\order{c},c^i]
\end{array}\]
